---
title: "BIOSTAT 650 Project"
author: "Jaehoon Kim (Group 19)"
date: "2024-11-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("sas7bdat")
library("corrplot")
library("gtsummary")
library("dplyr")
library("ggplot2")
library("data.table")
library("car")
library("NHANES")
library("olsrr")
```




```{r datasetup 1}
df = NHANES

```

Initial data exploration of covariates that had a relation to SexAge were difficult to perform a correlation plot due to being factors.
```{r EDA}
covariates = c("SexAge","Gender","HHIncome","Education","PhysActive","SameSex","AlcoholYear","RegularMarij","HardDrugs")
sapply(df[, covariates], is.factor)
#M = cor(df[, covariates])
#corrplot(M, method = 'number')
```

Running different multiple linear regressions, we found two models of interest after some exploratory data analysis with different covariates for which statistical significance persisted even after controlling for some social demographic covariates.
```{r model 1,2}
model <- lm(SexAge ~ RegularMarij+HardDrugs+RegularMarij*HardDrugs, df)
summary(model)

model <- lm(SexNumPartnLife ~ RegularMarij+HardDrugs+RegularMarij*HardDrugs, df)
summary(model)

```


SexAge is has a good distribution but SexNumPartnLife has extreme skenwness and is discrete count data. This requires a Poisson regression which is out side the scopre of this course. Created new variable using the duration, since first sexual activity where (Age - SexAge) since Age >= SexAge, and dividing by the number of sexual partners in life to see frequency of sexual activity. New variable was log transformed due to extreme skewness that violated normality assumption, which could be checked by QQPlot.

Due to extreme skewness,  we tried to find some observations that had implausible reported data that could been a typo or non serious answer. For instance, observations 8576 and 3416 reported to have had a first sexual activity at 9 with 360 and 500 sexual partners in life, respectively. Observations 4579 and 4580 reported to have had a first sexual activity at 10 and both reportedly had 700 sexual partners in life. Observations 4579 and 4580 reported to have had a first sexual activity at 10 and both reportedly had 700 sexual partners in life. We removed these outliers.
```{R New variable}
hist(df$SexAge, main= "First Age at which Sexual Activity Occured")
hist(df$SexNumPartYear, main = )
hist(df$SexNumPartnLife)

#Show observations with more than 300 sexual partners during lifetime
which(df$SexNumPartnLife > 300)
df[which(df$SexNumPartnLife > 300), c("Age", "SexAge", "SexNumPartnLife")]
df = df[-which(df$SexNumPartnLife > 300),]

#Before log transformation
df = mutate(df, AvgSexFreq = (Age-SexAge)/SexNumPartnLife)
hist(df$AvgSexFreq, main = "AvgSexFreq Before log transformation")
#After log transformation
df = mutate(df, AvgSexFreq = log((Age-SexAge)/SexNumPartnLife))
hist(df$AvgSexFreq, main = "AvgSexFreq After log transformation")
```

```{R New variable test}
#Remove negative infinity
df$AvgSexFreq[is.infinite(df$AvgSexFreq)] = NA
#unique(df$AvgSexFreq)

df$nPregnancies = is.factor(df$nPregnancies)
model <- lm(AvgSexFreq ~ SmokeNow+AlcoholYear+RegularMarij+HardDrugs+RegularMarij*HardDrugs+Age+Gender+HHIncome+Education+BMI+DiabetesAge+Depressed+LittleInterest+ PhysActive+SameSex, df)
summary(model)
model |>
  tbl_regression(intercept = TRUE)

#model <- lm(AvgSexFreq ~ #Gender+HHIncome+Education+PhysActive+SameSex+AlcoholYear+RegularMarij+HardDrugs+#RegularMarij*HardDrugs+RegularMarij*AlcoholYear, df)
#summary(model)

```

Using the sequential sum of squares we tested for each block of covariates at a significance level 0.05

```{r Sequential SS}
n = 50
aov = anova(model <- lm(AvgSexFreq ~ SmokeNow+AlcoholYear+RegularMarij+HardDrugs+RegularMarij*HardDrugs+Age+Gender+HHIncome+Education+BMI+DiabetesAge+Depressed+LittleInterest+PhysActive+SameSex, df))
aov
SSY = sum(aov$"Sum Sq")
SSQ = aov$"Sum Sq"
MSE = aov$"Mean Sq"[16]
ss1 = sum(SSQ[c(1:4, 15)])
print(ss1)
fstat1 = ss1/5/MSE
pval1 = 1-pf(q = fstat1, df1 = 5, df2 = n-16)
print(c(fstat1, pval1))
ss2 = sum(SSQ[5:8])
print(ss2)
fstat2 = ss2/4/MSE
pval2 = 1-pf(q = fstat2, df1 = 4, df2 = n-16)
print(c(fstat2, pval2))
ss3 = sum(SSQ[9:14])
print(ss3)
fstat3 = ss3/5/MSE
pval3 = 1-pf(q = fstat3, df1 = 5, df2 = n-16)
print(c(fstat3, pval3))
ss4 = sum(SSQ[14])
print(ss4)
fstat4 = ss3/1/MSE
pval4 = 1-pf(q = fstat4, df1 = 1, df2 = n-16)
print(c(fstat4, pval4))


```
(i) $\boldsymbol{\beta}_{substance} = (\beta_{SmokeNow}, \beta_{AlcoholYear},\beta_{RegularMarij}, \beta_{HardDrugs}, \beta_{RegularMarij*HardDrugs})^T$
(ii) $\boldsymbol{\beta}_{Demo} = (\beta_{Age}, \beta_{Gender}, \beta_{HHIncome}, \beta_{Education})^T$
(iii) $\boldsymbol{\beta}_{Health} = (\beta_{BMI},\beta_{DiabetesAges},\beta_{Depressed},\beta_{LittleInterest},\beta_{PhysActive})^T$
(iv) $\boldsymbol{\beta}_{SameSex} = (\beta_{SameSex})^T$


```{r tbl, message=FALSE, warnings=FALSE, echo=FALSE, results='asis'}
tbl <- c("
| Step | Tested Var. |  SS(Num.) | SS(Denom.) | Test Stat. | Dist. | p-value | Decision | Stopping Rule | Decision |
|------|----------|-----------|--------------|----------------|---------|-------------|-------------|------------|----------| 
|  $$I$$  |  $$\\boldsymbol{\\beta}_{Substance}$$ |   13.88444     |  26.9329 | 5.155204576 | $$F_{5,34}$$ | 0.001262146 | `Reject `| `Do not stop` | `Collect`|
|  $$II$$  |  $$\\boldsymbol{\\beta}_{Demo}$$ |   55.61473     |  26.9329 | 25.81174 | $$F_{4,34}$$ | 6.872507e-10 | `Reject `| `Do not stop` | `Collect`|
|  $$III$$  |  $$\\boldsymbol{\\beta}_{Health}$$ |   5.687399     |  26.9329 | 2.11169493 | $$F_{5,34}$$ | 0.08788892 | `Fail to Reject `| `Stop` | `Not Collect`|
|  $$IV$$  |  $$\\boldsymbol{\\beta}_{SameSex}$$ |   0.001708498     |  26.9329 | 10.55847467 | $$F_{1,34}$$ | 0.00260712 | `NA `| `NA` | `NA`|
")
cat(tbl, sep="")
```

```{r systematic differences of missing values}

library(ggplot2)
library(tidyr)
#Add new column based on missingness
covariates = c("AvgSexFreq", "SmokeNow","AlcoholYear", "RegularMarij", "HardDrugs", "Age", "Gender","HHIncome","Education")
sum(complete.cases(df[, covariates]))
df$missingness <- ifelse(complete.cases(df[, covariates]), "Missing", "Not Missing")


```
```{r missigness graphs}
library(gridExtra)
p1 = ggplot(data = df, mapping=aes(x=SmokeNow, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_fill_manual(values = c("gray", "red"))
p2 = ggplot(data = df, mapping=aes(x=AlcoholYear, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_fill_manual(values = c("gray", "red"))
p3 = ggplot(data = df, mapping=aes(x=RegularMarij, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_fill_manual(values = c("gray", "red"))
p4 = ggplot(data = df, mapping=aes(x=HardDrugs, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_fill_manual(values = c("gray", "red"))

grid.arrange(p1,p2,p3,p4, nrow=4)
```

```{r missingness plot 2}
p6 = ggplot(data = df, mapping=aes(x=Age, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_fill_manual(values = c("gray", "red"))
p7 = ggplot(data = df, mapping=aes(x=Gender, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_fill_manual(values = c("gray", "red"))
p8 = ggplot(data = df, mapping=aes(x=HHIncome, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
   scale_x_discrete(labels = c(1,2,3,4,5,6,7,8,9, 10, 11, 12, 13)) +
  scale_fill_manual(values = c("gray", "red"))
p9 = ggplot(data = df, mapping=aes(x=Education, fill=as.factor(missingness)))+
  geom_bar(stat="count")+
  scale_x_discrete(labels = c("<8th", "9-11th", "HS", "Some College", "Col Grad" )) +
  scale_fill_manual(values = c("gray", "red"))
grid.arrange(p6, p7, p8, p9, nrow = 4)
```


```{r QQplot}

m1 = lm(AvgSexFreq ~ SmokeNow+AlcoholYear+RegularMarij+HardDrugs+RegularMarij*HardDrugs+Age+Gender+HHIncome+Education, df)
m1.res = m1$residuals

car::qqPlot(m1.res)
```

```{r avPlots}

car::avPlots(m1)



```

```{r Residual Plots}


car::residualPlots(m1, type="response")
```

```{r multicollinearity}
#interactions(???)
nonintmodel <- lm(AvgSexFreq ~ SmokeNow+AlcoholYear+RegularMarij+Age+Gender+HHIncome+Education, df)
car::vif(nonintmodel,type = 'predictor')
```

```{r outliers}
model.deffits=dffits(m1)
model.CD = cooks.distance(m1)
model.deffits[which.max(model.deffits)]
model.CD[which.max(model.CD)]
n = nrow(df)
p = m1$rank
plot(m1, which = 4)
abline(h=4/n,lty=2)
df[c(2737, 3315, 8155),]

```

```{r plot for outlier}
ols_plot_resid_lev(m1)
df[c(475, 523, 944, 1473),]
```

```{r remove outliers}
df2 = df[-c(3315),]
m2 = lm(AvgSexFreq ~ SmokeNow+AlcoholYear+RegularMarij+HardDrugs+RegularMarij*HardDrugs+Age+Gender+HHIncome+Education, df2)
summary(m1)
summary(m2)
```

```{r rerun qqplot}
m2.res = m2$residuals

car::qqPlot(m2.res)
```